---
title: "Fragmentation by Chromosome"
author: "Haley Grant"
date: "3/27/2020"
output: 
  html_document:
    code_folding: "hide"
    toc: TRUE
    toc_float: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=F, warning=F}
library(tidyverse)
library(janitor)
library(dplyr)
library(readr)
library(ggplot2)
library(distrEx)
library(knitr)
library(kableExtra)
``` 

# Original 10 Regions 
The plots below show the distirbution of fragment lengths across the 10 regions (per chromosome) that Kamel identified. Details:

* The boxplots show the distribtution of fragment lengths across normal and cancer samples within each reagion of the specified chromosome.
* Because each sample has many reads per region, I use summary statistics for each distribution (within a sample, length of reads in region i of chromosome j). The summary statistics used in the plots are: mean, 10th percentile, median, and 90th percentile.
* The $\times$ marks show the estimated (average over samples) value of $1/\lambda$ for each region.
* The colors indicate cancer status. There are 4 healthy samples (first is a mix of many individuals, I believe) and about 40 cancer samples.


```{r message=F, warning=F}
d = "/Users/haleygrant/cancer_seek/data/shendure_data/chr_10region_dist"

chrom_files = paste("chr",seq(1,22),".csv",sep="")




```

```{r warning=F, message=F}

lvls = c()
for(i in 1:10){
  c = paste("r",i,"_cancer", sep = "")
  n = paste("r",i,"_normal",sep = "")
  lvls =rbind(lvls,c,n)
}



p =vector(mode = "list",length = 22)
cor_plt =vector(mode = "list",length = 22)
fit =vector(mode = "list",length = 22)
cors =vector(mode = "list",length = 22)



for(i in 1:length(chrom_files)){
  file = chrom_files[i]
  
  oldnames = paste("score",c("",paste(".",seq(1,9),sep="")),sep = "")
newnames = paste("r",seq(1,10),sep = "")

lambda_cancer = read_csv(paste("/Users/haleygrant/cancer_seek/data/Kamel_data/IntensityFragmentationNormalsCancers/LambdaDetectCancersFirst",i,".csv",sep = "")
                         )%>%
              rename_at(vars(oldnames), ~ newnames)%>%
              pivot_longer(3:12, names_to = "region", values_to = "score")%>%
              mutate(region = factor(region, levels = paste("r",seq(1,10) ,sep="")))
lambda_normal = read_csv(paste("/Users/haleygrant/cancer_seek/data/Kamel_data/IntensityFragmentationNormalsCancers/LambdaDetectNormalsFirst",i,".csv",sep = "")
                         )%>%
              rename_at(vars(oldnames), ~ newnames)%>%
              pivot_longer(3:12, names_to = "region", values_to = "score")%>%
              mutate(region = factor(region, levels = paste("r",seq(1,10) ,sep="")))



lambda = lambda_cancer%>%
  mutate(score = ifelse(score<=0,.001,score))%>%
  group_by(region)%>%
  summarise(lambda_inv = median(1/score, na.rm = T))%>%
  mutate(disease_status = "cancer")%>%
  bind_rows(lambda_normal%>%
              mutate(score= ifelse(score<=0,.001,score))%>%
  group_by(region)%>%
  summarise(lambda_inv = median(1/score, na.rm = T))%>%
  mutate(disease_status = "normal"))
  
  
 df = read_csv(paste(d,"/",file,sep = ""))%>%
   full_join(lambda, by = c("region","disease_status"))%>%
    mutate(region = factor(region, levels = paste("r",seq(1,10),sep = "")))%>%
    pivot_longer(4:13, names_to = "stat", values_to = "length")%>%
    mutate(region = paste(region, disease_status,sep="_"))%>%
    mutate(region = factor(region, levels =lvls))%>%
    filter(stat %in% c("mean_length","perc_10","perc_50","perc_90"))

 p[[i]]=df%>%
    ggplot(aes(x = region, y = length, color = disease_status))+
    geom_boxplot()+
   geom_point(aes(x = region, y = lambda_inv, color = disease_status), 
              shape = 4)+
    theme_bw()+
    scale_x_discrete(labels=gsub( "_.*", "", lvls))+
    scale_y_continuous(sec.axis = sec_axis(~., name ="1/lambda (mean)"))+
    theme(legend.position = "bottom", 
          axis.text.x = element_text(angle = 45, size = 8))+
    facet_wrap(.~stat, scales = "free_y")+
    labs(title = paste("Chromosome",i))

df_lm = df%>%
  group_by(stat, disease_status,region)%>%
  summarise(lambda_inv = ifelse(mean(lambda_inv, na.rm = T)<0,0,mean(lambda_inv,na.rm = T)),
            length= mean(length, na.rm = T))%>%
  filter(lambda_inv <125)%>%
  mutate(region = gsub("_.*",replacement = "", region))
  
cor_plt[[i]] = df_lm%>%
  ggplot(aes(x = lambda_inv, y = length))+
  geom_point(aes(color = disease_status))+
  theme_bw()+
  labs(x = "1/lambda")+
  geom_smooth(method = "lm", aes(color = disease_status))+
  facet_wrap(.~stat, scales = "free_y")





#correlation
c_mn = df_lm%>%
  na.omit()%>%
  ungroup()%>%
  filter(stat=="mean_length")%>%
  select(lambda_inv, length)%>%
  as.matrix()%>%
  cor(method = "spearman")
c_med = df_lm%>%
  na.omit()%>%
  ungroup()%>%
  filter(stat=="perc_50")%>%
  select(lambda_inv, length)%>%
  as.matrix()%>%
  cor(method = "spearman")
cors[[i]] = data.frame(mn_cor = round(c_mn[1,2],digits = 3), med_cor = round(c_med[1,2],digits = 3))
}






for(i in 1:22){
  print(p[[i]])
  print(cor_plt[[i]])

  
print(paste("Correlation between 1/lambda and mean length:", cors[[i]]))
}

data.frame(matrix(unlist(cors), nrow=length(cors), byrow=T))%>%
  rename(`Mean Correlation` = X1, `Median Correlation` = X2)%>%
  mutate(Chromosome = seq(1,22))%>%
  select("Chromosome", `Mean Correlation`,`Median Correlation`)%>%
  kable(align = "c")%>%
  kableExtra::kable_styling(bootstrap_options = c("striped","hover"), full_width = F)
```
# New regions from Kamel

```{r message=F, warning=F}
d_lasso = "/Users/haleygrant/cancer_seek/data/shendure_data/chr_lassoregions_dist"

chrom_files = paste("chr",seq(1,22),".csv",sep="")


```

```{r warning=F, message=F}


p =vector(mode = "list",length = 22)
cor_plt =vector(mode = "list",length = 22)
fit =vector(mode = "list",length = 22)
cors_lasso =vector(mode = "list",length = 22)
  
  all_chrom = data.frame(stat = NULL, disease_status = NULL, region = NULL, lambda_inv = NULL, length = NULL, sd_length = NULL, sd_lambda = NULL)
  
for(i in 1:length(chrom_files)){
  file = chrom_files[i]


lambda_cancer = read_csv(paste("/Users/haleygrant/cancer_seek/data/Kamel_data/LambdasCancerLasso/LambdaSeekCancersFirstLasso",i,".csv",sep = "")
                         )
nregions = ncol(lambda_cancer)-2
oldnames = paste("score",c("",paste(".",seq(1,nregions-1),sep="")),sep = "")
newnames = paste("r",seq(1,nregions),sep = "")

lambda_cancer = lambda_cancer%>%
              rename_at(vars(oldnames), ~ newnames)%>%
              pivot_longer(3:ncol(lambda_cancer), names_to = "region", values_to = "score")%>%
              mutate(region = factor(region, levels = paste("r",seq(1,20) ,sep="")))
lambda_normal = read_csv(paste("/Users/haleygrant/cancer_seek/data/Kamel_data/LamdasNormalLasso/LambdaDetectNormalsFirstLasso",i,".csv",sep = "")
                         )


lambda_normal = lambda_normal%>%
              rename_at(vars(oldnames), ~ newnames)%>%
              pivot_longer(3:ncol(lambda_normal), names_to = "region", values_to = "score")%>%
              mutate(region = factor(region, levels = paste("r",seq(1,20) ,sep="")))



lambda = lambda_cancer%>%
  mutate(score = ifelse(score<=0,.001,score))%>%
  group_by(region)%>%
  filter(score>.002)%>%
  summarise(sd_lambda = sd(1/score, na.rm = T), 
            lambda_inv = median(1/score, na.rm = T))%>%
  mutate(disease_status = "cancer")%>%
  bind_rows(lambda_normal%>%
              mutate(score = ifelse(score<=0,.001,score))%>%
  group_by(region)%>%
    filter(score>.002)%>%
  summarise(sd_lambda = sd(1/score, na.rm = T),
            lambda_inv = median(1/score, na.rm = T))%>%
  mutate(disease_status = "normal"))
  
 df = read_csv(paste(d_lasso,"/",file,sep = ""))%>%
   left_join(lambda, 
             by = c("region","disease_status"))%>%
    mutate(region = factor(region, levels = paste("r",seq(1,20),sep = "")))%>%
    pivot_longer(4:13, names_to = "stat", values_to = "length")%>%
    mutate(region = paste(region, disease_status,sep="_"))%>%
    filter(stat %in% c("mean_length","perc_10","perc_50","perc_90"))
  

  lvls = c()
for(j in 1:nregions){
  c = paste("r",j,"_cancer", sep = "")
  n = paste("r",j,"_normal",sep = "")
  lvls =rbind(lvls,c,n)
}

 p[[i]]=df%>%
    ggplot(aes(x = region, y = length, color = disease_status))+
    geom_boxplot()+
   geom_point(aes(x = region, y = lambda_inv, color = disease_status), 
              shape = 4)+
    theme_bw()+
    scale_x_discrete(labels=gsub("r*","",gsub( "_.*", "", lvls)))+
    scale_y_continuous(sec.axis = sec_axis(~., name ="1/lambda (mean)"))+
    theme(legend.position = "bottom", 
          axis.text.x = element_text(angle = 45, size = 8))+
    facet_wrap(.~stat, scales = "free_y")+
    labs(title = paste("Chromosome",i))

 
df_lm = df%>%
  group_by(stat, disease_status,region)%>%
  summarise(lambda_inv = ifelse(mean(lambda_inv, na.rm = T)<0,0,mean(lambda_inv,na.rm = T)),
                                sd_lambda = first(sd_lambda),
                                sd_length = sd(length, na.rm = T),
            length= mean(length, na.rm = T))%>%
  filter(lambda_inv<125)%>%
  mutate(region = gsub("_.*",replacement = "", region), chrom = i)

  all_chrom = bind_rows(all_chrom,df_lm)

  cor_plt[[i]] = df_lm%>%
  ggplot(aes(x = lambda_inv, y = length))+
  geom_point(aes(color = disease_status))+
  labs(x = "1/lambda")+
  theme_bw()+
  geom_smooth(method = "lm", aes(color = disease_status))+
  facet_wrap(.~stat, scales = "free_y")


#correlation
c_mn = df_lm%>%
  ungroup()%>%
  filter(stat=="mean_length")%>%
  select(lambda_inv, length)%>%
  as.matrix()%>%
  cor(method = "spearman")
c_med = df_lm%>%
  ungroup()%>%
  filter(stat=="perc_50")%>%
  select(lambda_inv, length)%>%
  as.matrix()%>%
  cor(method = "spearman")


cors_lasso[[i]] = data.frame(mn_cor = round(c_mn[1,2],digits = 3), med_cor = round(c_med[1,2],digits = 3))
}




for(i in 1:22){
  print(p[[i]])
  print(cor_plt[[i]])
}


data.frame(matrix(unlist(cors_lasso), nrow=length(cors_lasso), byrow=T))%>%
  rename(`Mean Correlation` = X1, `Median Correlation` = X2)%>%
  mutate(Chromosome = seq(1,22))%>%
  select("Chromosome", `Mean Correlation`,`Median Correlation`)%>%
  kable(align = "c")%>%
  kableExtra::kable_styling(bootstrap_options = c("striped","hover"), full_width = F)


all_chrom%>%
  mutate(chrom = as.character(chrom))%>%
  ggplot(aes(x = lambda_inv, y = length))+
  geom_point(size = 1, aes(color = chrom))+
  theme_bw()+
  facet_wrap(.~stat, scales = "free")

all_chrom%>%
  mutate(chrom = as.character(chrom))%>%
  ggplot(aes(x = sd_lambda, y = sd_length))+
  geom_point(size = 1, aes(color = chrom))+
  theme_bw()+
  facet_wrap(.~stat, scales = "free")
  
all_chrom%>%
  group_by(stat)%>%
  summarise(cor = cor(lambda_inv, length, method = "spearman"))

  
all_chrom%>%
  group_by(stat)%>%
  na.omit()%>%
  summarise(cor = cor(sd_length, sd_lambda, method = "spearman"))




print("Distribution of correlations for Lasso regions")
matrix(unlist(cors_lasso), nrow=length(cors_lasso), byrow=T)%>%summary()
print("Distribution of correlations for 10 regions")
matrix(unlist(cors), nrow=length(cors), byrow=T)%>%summary()
```

# Cancer - Normals (difference)


```{r warning=F, message=F}


p =vector(mode = "list",length = 22)
cors_diff =vector(mode = "list",length = 22)
  
all_chrom_diff = data.frame(stat = NULL, region = NULL, lambda_inv = NULL, length = NULL)
  
for(i in 1:length(chrom_files)){
  file = chrom_files[i]


lambda_cancer = read_csv(paste("/Users/haleygrant/cancer_seek/data/Kamel_data/LambdasCancerLasso/LambdaSeekCancersFirstLasso",i,".csv",sep = "")
                         )
nregions = ncol(lambda_cancer)-2
oldnames = paste("score",c("",paste(".",seq(1,nregions-1),sep="")),sep = "")
newnames = paste("r",seq(1,nregions),sep = "")

lambda_cancer = lambda_cancer%>%
              rename_at(vars(oldnames), ~ newnames)%>%
              pivot_longer(3:ncol(lambda_cancer), names_to = "region", values_to = "score")%>%
              mutate(region = factor(region, levels = paste("r",seq(1,20) ,sep="")))
lambda_normal = read_csv(paste("/Users/haleygrant/cancer_seek/data/Kamel_data/LamdasNormalLasso/LambdaDetectNormalsFirstLasso",i,".csv",sep = "")
                         )


lambda_normal = lambda_normal%>%
              rename_at(vars(oldnames), ~ newnames)%>%
              pivot_longer(3:ncol(lambda_normal), names_to = "region", values_to = "score")%>%
              mutate(region = factor(region, levels = paste("r",seq(1,20) ,sep="")))



lambda = lambda_cancer%>%
  mutate(score = ifelse(score<=0,.001,score))%>%
  group_by(region)%>%
  summarise(lambda_inv = median(1/score, na.rm = T))%>%
  mutate(disease_status = "cancer")%>%
  bind_rows(lambda_normal%>%
              mutate(score = ifelse(score<=0,.001,score))%>%
  group_by(region)%>%
  summarise(lambda_inv = median(1/score, na.rm = T))%>%
  mutate(disease_status = "normal"))
  
 df = read_csv(paste(d_lasso,"/",file,sep = ""))%>%
   left_join(lambda, 
             by = c("region","disease_status"))%>%
    mutate(region = factor(region, levels = paste("r",seq(1,20),sep = "")))%>%
    pivot_longer(4:13, names_to = "stat", values_to = "length")%>%
    filter(stat %in% c("mean_length","perc_10","perc_50","perc_90"))%>%
   group_by(disease_status, region, stat)%>%
   summarise( length = mean(length), lambda_inv = mean(lambda_inv))%>%
   pivot_wider(names_from = disease_status, 
               values_from = c("length","lambda_inv"))%>%
   group_by(region, stat)%>%
   filter(lambda_inv_normal < 125 &lambda_inv_cancer <125 )%>%
   mutate(length= length_normal-length_cancer,
             lambda_inv = lambda_inv_normal - lambda_inv_cancer)%>%
   select(region, stat, length, lambda_inv)
   
 all_chrom_diff = bind_rows(all_chrom_diff,df)
  
 p[[i]]=df%>%
    ggplot()+
    geom_point(aes(y = length, x = lambda_inv))+
    theme_bw()+
    theme(legend.position = "none", 
          axis.text.x = element_text(angle = 45, size = 8))+
    facet_wrap(.~stat, scales = "free")+
    labs(title = paste("Chromosome",i))

 



#correlation
c_mn = df%>%
  na.omit()%>%
  ungroup()%>%
  filter(stat=="mean_length")%>%
  select(lambda_inv, length)%>%
  as.matrix()%>%
  cor(method = "spearman")

c_med = df%>%
  na.omit()%>%
  ungroup()%>%
  filter(stat=="perc_50")%>%
  select(lambda_inv, length)%>%
  as.matrix()%>%
  cor(method = "spearman")
cors_diff[[i]] = data.frame(mn_cor = round(c_mn[1,2],digits = 3), med_cor = round(c_med[1,2],digits = 3))
}




for(i in 1:22){
  print(p[[i]])
}


data.frame(matrix(unlist(cors_diff), nrow=length(cors_diff), byrow=T))%>%
  rename(`Mean Correlation` = X1, `Median Correlation` = X2)%>%
  mutate(Chromosome = seq(1,22))%>%
  select("Chromosome", `Mean Correlation`,`Median Correlation`)%>%
  kable(align = "c")%>%
  kableExtra::kable_styling(bootstrap_options = c("striped","hover"), full_width = F)


all_chrom_diff%>%
  group_by(stat)%>%
  summarise(cor = cor(lambda_inv, length, method = "spearman"))
  
```

# Region widths

```{r message=FALSE, warning=FALSE}
widths = data.frame(chrom = NULL, region = NULL, width = NULL)
for(j in 1:22){

regions_dir = "/Users/haleygrant/cancer_seek/data/Kamel_data/AmpliconLassoRegions/"
regions = read_csv(paste(regions_dir,"RegionsOfAmpliconsLasso",j,".csv",sep = ""))


region_boundaries = regions%>%
  clean_names()%>%
  mutate(region = paste("r",region,sep = ""),
         region = factor(region, levels = c(paste("r",seq(1,20),sep=""))))%>%
  group_by(region)%>%
  summarise(min = min(end, na.rm = T),
            max = max(end, na.rm = T))%>%
  mutate(width = max - min, chrom = j)
  

widths = bind_rows(widths, region_boundaries%>%select(chrom, region, width))

}
widths%>%
  mutate(chrom = factor(chrom, levels = as.character(seq(1:22))))%>%
  mutate(chrom_grp = ifelse(as.numeric(chrom)<6, "Chrom 1-5", ifelse(as.numeric(chrom)<10, "Chrom 6-10", ifelse(as.numeric(chrom)<17, "Chrom 11-16","Chrom 17-22"))))%>%
  ggplot(aes(x = width, color = chrom, group = chrom))+
  geom_density(aes(color = chrom))+
  theme_bw()+
  labs(title = "Width of Regions")+
  facet_wrap(.~chrom_grp, scales = "free_y")



```